---
title: "Vignette Title"
author: "Emanuele Cordano"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo=TRUE,
  warning = FALSE
)
###knitr::opts_chunk$set(echo = TRUE)
```


<!-- --- -->
<!-- title: "Untitled" -->
<!-- output: github_document -->
<!-- bibliography: bibliography.bib -->

<!-- --- -->

## lmomIDFGitHub Documents


Intensity-duration-frequency (IDF) curves usefully quantify extreme precipitation over various duration, e.g. from 1 to 360 hours,  and return periods for engineering design, e.g. the interest in the time concentration or time-structure of the rainfall.  (e.g. @Courty2019,@Sun2019)
Generally  a mathematical function that relates the rainfall intensity with its duration and frequency of occurrence, as follows (@Koutsoyiannis1998): 

LOREM ISPUM 
Recently, @Heidari2020 extend also the concepts of IDF to drought. This aspect can be used in this package but requires further investigations. 

## Including Code

Here is an example of use of the packege with GSOD precipitation data through GSODR package (@Sparks2017): 

```{r gsod}

library(GSODR)
library(dplyr)
library(magrittr)
years <- 1937:2020
gsod_rds <- "/home/ecor/local/rpackages/jrc/lmomIDF/inst/ext_data/gsod.rds"
cond <- file.exists(gsod_rds) 
if (!cond) {
  gsod <- get_GSOD(years=years,station="623180-99999") ##ALEXANDRIA INTL EG 623180-99999
  saveRDS(gsod,file=gsod_rds)  
} else {
  gsod <- readRDS(file=gsod_rds)
}

#####
#####

prec <- gsod %>% select(YEARMODA,PRCP,PRCP_ATTRIBUTES) %>%
 mutate(YEARMODA=as.Date(YEARMODA,format="%Y-%m-%d")) 

knitr::kable(rbind(head(prec),tail(prec)))

### Dataset preparation 
## insertion of rows so that all instants are equidistant (i.e. there is a constant time step) 
## inserting any NA values of precipitation intensity


dds <- range(prec$YEARMODA)
## See GSODR documentation
yymmdds <- seq(from=dds[1],to=dds[2],by="day")
prec <- data.table::data.table(YEARMODA=yymmdds) %>% full_join(prec)



```



## Precipitation Time Series

Time series can be visualized as follows: 

```{r gsod_zoo, fig.width=7, paged.print=FALSE}

library(zoo)
library(dygraphs)

time <- prec$YEARMODA
precz <- prec$PRCP %>% as.zoo()
index(precz) <- time
main="DAILY PRECIPITATION Vs TIME"
ylab="precipitation [mm/hr]"
xlab="time"

dd <- dygraph(precz,main=main,ylab=ylab,xlab=xlab) %>% dyRangeSelector()
dd


```

## Annual Maxima for fixed duration 

Annual maxima are extracted:
```{r ann.max,fig.width=7, paged.print=FALSE}

library(lmomIDF)

ann.maxima <- annual.agg(x=prec$PRCP,time=prec$YEARMODA,aggr.fun=max,dd=1:5)
### values cannot be -Inf or +Inf
ann.maxima$aggr[ann.maxima$aggr==-Inf] <- NA
###
str(ann.maxima)
###
ann.maxima.z <- ann.maxima %>% reshape2::dcast(index ~ dd)
time <- as.numeric(ann.maxima.z$index)
ann.maxima.z <- ann.maxima.z %>% select(-index) %>% as.zoo()
index(ann.maxima.z) <- time
main="MAXIMUM PRECIPITATION INTENSITY Vs TIME"
ylab="precipitation intensity [mm/hr]"
xlab="time"

dda <- dygraph(ann.maxima.z,main=main,ylab=ylab,xlab=xlab) %>% dyRangeSelector()
dda
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## References 


```{r generateBibliography,echo=FALSE,eval=TRUE,message=FALSE,warning=FALSE,print=FALSE,results="hide"}

require(knitcitations)
cleanbib()
options(citation_format="pandoc")
read.bibtex(file = "bibliography.bib")


```
















<!-- ## START VIGNETTE TEMPLATE -->

<!-- Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format: -->

<!-- - Never uses retina figures -->
<!-- - Has a smaller default figure size -->
<!-- - Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style -->

<!-- ## Vignette Info -->

<!-- Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette. -->

<!-- ## Styles -->

<!-- The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows: -->

<!--     output:  -->
<!--       rmarkdown::html_vignette: -->
<!--         css: mystyles.css -->

<!-- ## Figures -->

<!-- The figure sizes have been customised so that you can easily put two images side-by-side.  -->

<!-- ```{r, fig.show='hold'} -->
<!-- plot(1:10) -->
<!-- plot(10:1) -->
<!-- ``` -->

<!-- You can enable figure captions by `fig_caption: yes` in YAML: -->

<!--     output: -->
<!--       rmarkdown::html_vignette: -->
<!--         fig_caption: yes -->

<!-- Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**. -->

<!-- ## More Examples -->

<!-- You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`. -->

<!-- ```{r, echo=FALSE, results='asis'} -->
<!-- knitr::kable(head(mtcars, 10)) -->
<!-- ``` -->

<!-- Also a quote using `>`: -->

<!-- > "He who gives up [code] safety for [code] speed deserves neither." -->
<!-- ([via](https://twitter.com/hadleywickham/status/504368538874703872)) -->
